A Comparison of Neural Network Models for Water Quality Forecasting

### 2024-06-12 manuscript/manuscript.Rmd
Making ground on the disc section

In Table \ref{}, we present the mean CRPS and RMSE scores for the respective target variables. For Dissolved Oxygen (DO) and Water Temperature (WT), there are similar results for model performance: the neural network models outperform the AutoTheta, Naive Persistence and Climatology model with few exceptions; and the naive ensemble model is the best performing model with respect to both CRPS and RMSE. Yet, with Chlorophyll-A, there are much different results: while the neural network models still generally outperform the reference models in CRPS, the Naive Ensemble model is no longer the top performing model in either CRPS or RMSE. BlockRNN attains the best performance with respect to CRPS and the Naive Persistence model attains the best mean RMSE.

Intuition for why we see the results in the scoring tables can be gained by looking at individual forecasts shown in Figure \ref{}.

@@

Over the last decade, ML has gained significant prominence after a spate of high profile successes like beating the master champion of Go, accurately predicting protein structures and the advent of large language models like ChatGPT \cite{alphago, alphafold, chatgpt}. During this time, researchers have developed many ML models for time series forecasting with promises of significant accuracy improvements over previous methods \cite{makridakis}. For this study, we selected 8 ML models that include s

Yet, these claims of improvement tend to have major limitations, notably that the studies examine a limited amount of time series and they do not compare to naive benchmarks \cite{makridakis}.

Models tend to struggle to accurately forecast chlorophyll during bloom conditions \cite{woelmer}. Accurately predicting timing and magnitude of bloom events is not surprising

>>>>

In Table \ref{}, we present the mean CRPS and RMSE scores for the respective target variables. For Dissolved Oxygen (DO) and Water Temperature (WT), the models perform similarly: the neural network models outperform the AutoTheta, Naive Persistence and Climatology model with few exceptions; and the naive ensemble model is the best performing model with respect to both CRPS and RMSE. Yet, with Chlorophyll-A, there are different patterns in model performance: while the neural network models still generally outperform the reference models in CRPS, the Naive Ensemble model is no longer the top performing model in either CRPS or RMSE; and, BlockRNN and the Naive Persistence model attain the best performance with respect to CRPS and RMSE, respectively.

By examining some of the individual forecasts (Figure \ref{}) used for the evaluations, it is possible to gain intuition for why the neural network models perform well across the target variables. The AutoTheta model produces forecasts that look similar to a linear regression based upon recently observed values. These relatively simple forecasts perform well for many evaluation intervals, but there a few cases when AutoTheta fails significantly, which negatively impacts the model's overall performance. For instance, after a peak of dissolved oxygen in the winter of 2024, the AutoTheta model forecasts that DO will continue to increase which is opposed to the trend that DO peaks in the winter and declines through the spring. Similarly with Chlorophyll-A, the AutoTheta model wrongly extrapolates that a spike in Chlorophyll-A will lead to even higher Chlorophyll-A concentrations instead of an immediate reversion to the non-bloom state.

Conversely, the neural network models are able to learn from historical trends. For instance, at the beginning of 2024, the neural network models have learned from the training data that DO peaks in the winter and declines through the spring, so the neural networks correctly predicts that there will be a decline in dissolved oxygen during this month. With Chlorophyll-A, however, we see that the neural network models fail to predict any of the spikes in concentration which originate from algae blooms; instead, the neural networks take a conservative approach, only predicting that the chla concentration will remain at non-bloom levels throughout the year.

These general patterns in model performance can also be seen in the skill score plots displayed in Figures \ref{}. The AutoTheta model tends to have longer tails, indicating that the AutoTheta forecasts are more likely to perform very well or very badly. Meanwhile, the neural network models generally have fewer forecasts that underperform relative to AutoTheta and less forecasts that are outlier outperformers, but the distributions of NN skill scores are translated towards outperforming. In these plots, we display scores according to water body type, and we did not see any significant differences in performance across these categories. This may be due to an underreporting of scores for lakes and non-wadeable rivers relative to the number of scores found for wadeable streams.

When examining the performance of the models within a forecast horizon, additional nuances emerge, confusing the perspective that neural network models are the best performing model class. In Figure \ref{}, the AutoTheta model is consistently the best performing model for short time horizons across target variables, but by the end of the 30-day horizon, the AutoTheta model is the worst perfoming model universally. Meanwhile the neural network models underperform AutoTheta during the early stages of the forecast window (generally, when t < 5), but their forecast skill declines less rapidly than the skill score of AutoTheta. So while, the neural networks generally outperform AutoTheta and the null models according to coarse scoring metrics like mean CRPS and RMSE, AutoTheta is the best performer in short horizons.

@@

Over the last decade, machine learning has gained significant prominence after a spate of high profile successes based on neural network models, some of these successes include beating the master champion of Go, accurately predicting protein structures and the advent of large language models like ChatGPT \cite{alphago, alphafold, chatgpt}. During this time, researchers in computer science have developed new neural network models for time series forecasting with promises of significant accuracy improvements over previous methods \cite{makridakis}. Simultaneously during the last 10 years, there has been a growing focus in limnology towards using neural networks for water quality forecasting, but this focus has primarily centered on older neural network architectures and there has been little exploration of newer methods. This study aims to address this research gap by performing a comprehensive comparison of neural network models, including models taken from more recent research in computer science.

Instead of identifying that there is some new neural network model that performs exceptionally well, this study has affirmed the well-established forecasting axiom that ensemble forecasts are often more accurate and robust than forecasts from single models. We found that the Naive Ensemble model which aggregated the forecasts from the neural network models was the best performing model overall, evincing all the other models on the target variables of dissolved oxygen and water temperature, while also performing well for chlorophyll-a (attaining the second and third best CRPS and RMSE, respectively). This result affirms recent work from \cite{freya} that established multi-model ensembles as advantageous over individual models for water temperature forecasting.

Yet, as is often the case when judging whether one method is superior to another, we have found that subtle changes to our criteria may have produced radically different results. For instance, it is possible that if we changed the forecast horizon to 5 days, then the AutoTheta model may have been the best performing model as the AutoTheta model outperformed all the other models in short time horizons (see Figure \ref{}). But, it is important to qualify that changing our criteria to consider shorter term forecasts may result in the neural network models learning different behavior during training. How these models perform with different forecast horizons warrants future research.

The neural network models struggled to forecast Chlorophyll-A, a result that is not surprising given that forecasting algae blooms has been well established as a challenging prediction problem in limnology \cite{}. The neural network models displayed a tendency to take a conservative strategy, regularly predicting that the chlorophyll-a concentration would remain in a non-bloom state. It is possible that the neural network models settled on this conservative behavior as they were not provided enough information to predict blooms -- the neural networks were not provided with some of the typical covariates used in process-based chlorophyll-a models like Nitrogren and Phosphorous concentrations as well as measures of photosynthetic active radiation. And, it is likely that this conservative strategy is an artifact of the length of the forecast horizon. Blooms are stochastic events, characterized by rapid fluctuations in chlorophyll-a. Predicting the exact time of such an event in a long term horizon will be more difficult than predicting the onset of a bloom in the near-term \cite{}.

We hope that this comparative study will encourage more research that investigates neural networks for water quality forecasting. In place of the one research question that we attempted to answer, "How do neural network models perform for water quality forecasting?", many new questions and ideas have arose. Outside of exploring how the performance of neural networks will vary with forecast horizon, which was mentionned previously, we think that some other promising directions for exploration include cross-learning and hybrid models.

Provided that there are large limnological datasets that currently exist, and there will be an increasing amount of limnological data coming online, neural network models offer significant advantages over process-based and statistical models as they will be able to learn directly from the data. This data-driven approach is advantageous in a few ways: as NN models can learn directly from the data, these models will not be limited by the assumptions inherent to process-based and statistical models. Moreover, these models should improve as
